Notes

### 2022-09-21 notes/notes.tex
simplify expression for gradient

\section{Notation}
\begin{itemize}
    \item Average emission per source: $x_1,\dots,x_m$ where $m = 200$.
    \item Measured concentrations $y_1,\dots,y_n$ where $n = 4636$. These are consecutive measurements for several measurement stations. 
    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta - 1$ days ago to the observed concentration $y_i$.
    \item Scaling factors $s_1,\dots,s_m$ (in \texttt{scalings.csv}). These are put in a diagonal matrix $S$.
    \item $\vb{1}_q = [1\,\, \dots\,\, 1]^T \in \mathbb{R}^q$ is the constant vector consisting of only ones.
    \item $\otimes$ is the Kronecker product. In particular,
        $$
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{p}}
        \otimes 
        \underbrace{
        \begin{bmatrix}
            1 \\
            \vdots \\
            1
        \end{bmatrix}
        }_{\in \mathbb{R}^{q}}
        =
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_1 \\
            a_2 \\
            \vdots \\
            a_2 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{pq}}
        .$$
        A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly. See, e.g. \cite[Section 12.3]{Golub2013}.
\end{itemize}

@@

\subsection{Fitting the model}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
.$$
\begin{proposition}
    Let $\hat{z}$ be the vector $[\hat{z}_1\,\,\dots\,\,\hat{z}_n]^T$ where 
    $$
    \hat{z}_i = \begin{cases}
        \frac{\log \hat{y}_i}{\hat{y}_i} & \text{if observation $i$ is a detection} \\
        0 & \text{if observation $i$ is a non-detection} 
    \end{cases}
    .$$
    Then the gradient of $f(W)$ is the diagonal matrix
    $$
    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left(M_{shift}^T \hat{z} \tilde{x}^T\right) + \lambda W^{-1} \log W
    $$
    where $\mathrm{Diag}$ is the map that sets the off-diagonal elements of a matrix to zero.
\end{proposition}
\begin{proof}

Parameterise $W$ linearly as $\mathrm{vec}\, W(t) = U t$ for some unitary $U$.
Define $$
G(t) :=
P_{det} \hat{y}_i
=
P_{det} M_{shift} W(t) \tilde{x}
\quad\text{and}\quad
F(t) := \log.(G(t))
.$$
Then 
$$
G(t) = P_{det} M_{shift} W(t) \tilde{x} = \underbrace{(\tilde{x}^T \otimes P_{det}M_{shift}) U}_A t
.$$
Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
$$
\frac{\partial}{\partial t} F(t) =
\frac{\partial}{\partial t} \log.(G(t)) = \mathrm{diag}\,(G(t))^{-1} A
.$$
Write $z := P_{det} \hat{z}$.
We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(t)$.
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right) = \left(\frac{\partial F}{\partial t}\right)^T F(t)
&= A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
&= A^T z \\
&= U^T (\tilde{x} \otimes (P_{det} M_{shift})^T) z \\
&= U^T (P_{det} M_{shift})^T z \tilde{x}^T \\
&= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}^T \\
&= U^T M_{shift}^T z \tilde{x}^T
\end{align*}
where the last equality follows from the fact that $P_{det}^T P_{det}$ is the map that sets the rows corresponding to the non-detection events to zero. 
Since $U$ is unitary, we have $$
\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
U \mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
UU^T (P_{det} M_{shift})^T z \tilde{x}^T
$$
The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices, which is equivalent to $\mathrm{Diag}$.

>>>>

\section{Notation}
\begin{itemize}
    \item Average emission per source: $x_1,\dots,x_m$ where $m = 200$.
    \item Measured concentrations $y_1,\dots,y_n$ where $n = 4636$. These are consecutive measurements for several measurement stations. 
    \item $M_{i,(j,\Delta)}$ is a block in the sensitivity matrix $M$ that represents the contribution of $x_j$ from $\Delta - 1$ days ago to the observed concentration $y_i$.
    \item Scaling factors $s_1,\dots,s_m$ (in \texttt{scalings.csv}). These are put in a diagonal matrix $S$.
    \item $\vb{1}_q = [1\,\, \dots\,\, 1]^T \in \mathbb{R}^q$ is the constant vector consisting of only ones.
    \item $\odot$ is the Hadamard (i.e.\, elementwise) product, given by $(AB)_{ij} = A_{ij} B_{ij}$.
    \item $\mathrm{diag}$ is the canonical identification of $p$-vectors and $p \times p$ diagonal matrices, i.e.,
    $$
    \mathrm{diag}\left(
    \begin{bmatrix}
        a_1 \\ 
        \vdots \\
        a_p
    \end{bmatrix}
    \right)
    =
    \begin{bmatrix}
        a_1 & 0 & \dots & 0 \\
        0 & a_2 & \dots & 0 \\
        \vdots & \vdots & \ddots & 0 \\
        0 & 0 & \cdots & a_p
    \end{bmatrix}
    .$$
    \item $\otimes$ is the Kronecker product. In particular,
        $$
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{p}}
        \otimes 
        \underbrace{
        \begin{bmatrix}
            1 \\
            \vdots \\
            1
        \end{bmatrix}
        }_{\in \mathbb{R}^{q}}
        =
        \underbrace{
        \begin{bmatrix}
            a_1 \\
            \vdots \\
            a_1 \\
            a_2 \\
            \vdots \\
            a_2 \\
            \vdots \\
            a_p
        \end{bmatrix}
        }_{\in \mathbb{R}^{pq}}
        .$$
        A standard rule of thumb in numerical linear algebra is that the Kronecker product serves a mostly theoretical purpose and is rarely calculated explicitly. See, e.g. \cite[Section 12.3]{Golub2013}.
\end{itemize}

@@

\subsection{Fitting the model}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
.$$
\begin{proposition}
    Let $\hat{z}$ be the vector $[\hat{z}_1\,\,\dots\,\,\hat{z}_n]^T$ where 
    $$
    \hat{z}_i = \begin{cases}
        \frac{\log \hat{y}_i}{\hat{y}_i} & \text{if observation $i$ is a detection} \\
        0 & \text{if observation $i$ is a non-detection} 
    \end{cases}
    .$$
    Then the gradient of $f(W)$ is the diagonal matrix
    $$
    \mathrm{grad}_{W} f(W) = \mathrm{diag}\left(\left(M_{shift}^T \hat{z}\right) \odot \tilde{x}\right) + \lambda W^{-1} \log W
    .$$
\end{proposition}
\begin{proof}

Parameterise $W$ linearly as $\mathrm{vec}\, W(t) = U t$ for some unitary $U$.
Define $$
G(t) :=
P_{det} \hat{y}_i
=
P_{det} M_{shift} W(t) \tilde{x}
\quad\text{and}\quad
F(t) := \log.(G(t))
.$$
Then 
$$
G(t) = P_{det} M_{shift} W(t) \tilde{x} = \underbrace{(\tilde{x}^T \otimes P_{det}M_{shift}) U}_A t
.$$
Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
$$
\frac{\partial}{\partial t} F(t) =
\frac{\partial}{\partial t} \log.(G(t)) = \mathrm{diag}\,(G(t))^{-1} A
.$$
Write $z := P_{det} \hat{z}$.
We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(t)$.
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right) = \left(\frac{\partial F}{\partial t}\right)^T F(t)
&= A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
&= A^T z \\
&= U^T (\tilde{x} \otimes (P_{det} M_{shift})^T) z \\
&= U^T (P_{det} M_{shift})^T z \tilde{x}^T \\
&= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}^T \\
&= U^T M_{shift}^T \hat{z} \tilde{x}^T
\end{align*}
where the last equality follows from the fact that $P_{det}^T P_{det}$ is the map that sets the rows corresponding to the non-detection events to zero. 
Since $U$ is unitary, we have $$
\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
U \mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
UU^T M_{shift}^T \hat{z} \tilde{x}^T
$$
The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices. This orthogonal projection simply sets the off-diagonal elements of its argument to zero. Since the diagonal of $M_{shift}^T \hat{z} \tilde{x}^T$ is $(M_{shift}^T \hat{z}) \odot \tilde{x}$, we have 
$$
\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
\mathrm{diag}((M_{shift}^T \hat{z}) \odot \tilde{x})
.$$
